---
title: "Data Preparation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Preparation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE)
```

This vignette processes the raw Afrotropical frugivory data to create the key
data structures needed for model fitting. It must be run before any subsequent
analysis vignettes.

```{r paths}
library(COILplus)
library(tidyverse)
library(foreach)
library(doParallel)
library(abind)

base_path      <- tools::R_user_dir("COILplus", which = "data")
raw_path       <- file.path(base_path, "raw")
processed_path <- file.path(base_path, "processed")

dir.create(raw_path,       recursive = TRUE, showWarnings = FALSE)
dir.create(processed_path, recursive = TRUE, showWarnings = FALSE)

save_files <- TRUE
```

## Load raw data

First we download sample data from the Zenodo record: https://zenodo.org/records/20357089/
```{r download}
zenodo_base <- "https://zenodo.org/records/20357089/files/"
files <- c("Cu_phylo.dat", "Cv_phylo.dat", "frug_generic.csv",
           "Obs_W.dat", "Obs_X.dat", "p_taxa_generic.csv",
           "v_taxa_generic.csv", "Readme.md")

for (f in files) {
  dest <- file.path(raw_path, f)
  if (!file.exists(dest)) {
    download.file(paste0(zenodo_base, f, "?download=1"), 
                  destfile = dest, mode = "wb")
  }
}

```

```{r load}
frug <- read.csv(file.path(raw_path, "frug_generic.csv"))
```

## Part A: Exploratory summary

First we explore the basic data structure and note that individual studies may contain multiple different sampling schemes (e.g., observations and separate locations), each of which is given its own unique identifier. 

```{r eda}
apply(frug, 2, function(x) sum(is.na(x)))
table(frug$Zone)
table(frug$Country)
table(frug$Focus)
nrow(frug)

frug_unique <- frug %>%
  group_by(Animal_Species_Corrected, Plant_Species_Corrected) %>%
  dplyr::select(Animal_Species_Corrected, Plant_Species_Corrected) %>%
  unique()
nrow(frug_unique)

v.names <- unique(frug$Animal_Species_Corrected)
p.names <- unique(frug$Plant_Species_Corrected)
s.names <- unique(frug$Study_ID)
sp.names <- unique(frug$Study_ID_Plus)

nV <- length(v.names)
nP <- length(p.names)
nS <- length(s.names)
nSP <- length(sp.names)
```

## Part B: Create the adjacency array

Convert the frugivory data from long format to an adjacency array with rows corresponding to frugivore species, columns corresponding to plant species, and slices (dimension three) corresponding to sampling units. 
```{r adjacency}
frug_new <- frug %>%
  dplyr::select(Animal_Species_Corrected, Plant_Species_Corrected, Study_ID, Study_ID_Plus)
frug_new$Interaction <- 1

Adj_matrix_new <- list()
for (ss in 1:nSP) {
  Adj <- frug_new %>%
    filter(Study_ID_Plus == sp.names[ss]) %>%
    dplyr::select(Animal_Species_Corrected, Plant_Species_Corrected, Interaction) %>%
    distinct() %>%
    pivot_wider(names_from = Plant_Species_Corrected, values_from = Interaction)
  if (dim(Adj)[2] == 2) {
    this.v <- Adj$Animal_Species_Corrected
    this.p <- colnames(Adj)[2]
    Adj_df <- as.data.frame(Adj)[, -1]
    Adj_matrix <- as.matrix(Adj_df)
    dimnames(Adj_matrix) <- list(this.v, this.p)
  } else {
    Adj_df <- as.data.frame(Adj)
    rownames(Adj_df) <- Adj_df$Animal_Species_Corrected
    Adj_df <- Adj_df[, -1]
    Adj_df[is.na(Adj_df)] <- 0
    Adj_matrix <- as.matrix(Adj_df)
  }
  s.id <- sp.names[ss]
  Adj_matrix_new[[s.id]] <- Adj_matrix
}

acomb <- function(...) abind(..., along = 3)
registerDoParallel(10)

A_obs <- foreach(i = 1:nSP, .combine = "acomb") %dopar% {
  A <- Adj_matrix_new[[i]]
  wh.v <- sapply(dimnames(A)[[1]], function(x) which(x == v.names))
  wh.p <- sapply(dimnames(A)[[2]], function(x) which(x == p.names))
  A.full <- matrix(0, nrow = nV, ncol = nP)
  A.full[wh.v, wh.p] <- A
  A.full
}
dimnames(A_obs) <- list(v.names, p.names, sp.names)
```

## Part D: Create occurrence matrices

Occurrence probabilities are assigned based on geographic overlap:
1 = same study, 0.75 = same site, 0.5 = same country, 0.25 = same zone, 0 = none.

```{r occurrence}
O_P <- matrix(0, nrow = nP, ncol = nSP)
rownames(O_P) <- p.names
O_V <- matrix(0, nrow = nV, ncol = nSP)
rownames(O_V) <- v.names
colnames(O_P) <- colnames(O_V) <- sp.names

for (ss in 1:nSP) {
  IDplus <- sp.names[ss]
  study_site    <- unique(frug[frug$Study_ID_Plus == IDplus, "Study_Site"])
  study_zone    <- unique(frug[frug$Study_ID_Plus == IDplus, "Zone"])
  study_country <- unique(frug[frug$Study_ID_Plus == IDplus, "Country"])

  study_site_dta <- frug[which(frug$Study_ID_Plus == IDplus), ]
  site_dta    <- frug[which(frug$Study_Site == study_site & !is.na(frug$Study_Site)), ]
  country_dta <- frug[which(frug$Country == study_country & !is.na(frug$Country)), ]
  zone_dta    <- frug[which(frug$Zone == study_zone & !is.na(frug$Zone)), ]

  for (p in 1:nP) {
    this.plant <- p.names[p]
    if (length(study_site_dta$Plant_Species_Corrected[study_site_dta$Plant_Species_Corrected == this.plant]) > 0) {
      O_P[p, ss] <- 1
    } else if (length(site_dta$Plant_Species_Corrected[site_dta$Plant_Species_Corrected == this.plant]) > 0) {
      O_P[p, ss] <- 0.75
    } else if (length(country_dta$Plant_Species_Corrected[country_dta$Plant_Species_Corrected == this.plant]) > 0) {
      O_P[p, ss] <- 0.5
    } else if (length(zone_dta$Plant_Species_Corrected[zone_dta$Plant_Species_Corrected == this.plant]) > 0) {
      O_P[p, ss] <- 0.25
    }
  }
  for (v in 1:nV) {
    this.animal <- v.names[v]
    if (length(study_site_dta$Animal_Species_Corrected[study_site_dta$Animal_Species_Corrected == this.animal]) > 0) {
      O_V[v, ss] <- 1
    } else if (length(site_dta$Animal_Species_Corrected[site_dta$Animal_Species_Corrected == this.animal]) > 0) {
      O_V[v, ss] <- 0.75
    } else if (length(country_dta$Animal_Species_Corrected[country_dta$Animal_Species_Corrected == this.animal]) > 0) {
      O_V[v, ss] <- 0.5
    } else if (length(zone_dta$Animal_Species_Corrected[zone_dta$Animal_Species_Corrected == this.animal]) > 0) {
      O_V[v, ss] <- 0.25
    }
  }
}
```

## Part E: Create the focus array

The focus array `F_obs` encodes which species pairs were "in scope" for observation in each study, based on the study's sampling method. This is necessary because different study designs have different observation biases:

- **Animal-focused studies** record any interaction involving a focal animal species, so all plants are considered in scope for any animal that was observed at all.
- **Plant-focused studies** record any interaction involving a focal plant species, so all animals are considered in scope for any plant that was observed at all.
- **Both-limited or unknown studies** are assumed to only report interactions that were actually observed — the observed interaction matrix itself defines what was in scope.
- **Network-wide studies** (all other methods) observed all possible animal-plant combinations, so all pairs are in scope.


```{r focus}
F_obs <- array(0, dim = c(nV, nP, nSP))
dimnames(F_obs) <- list(v.names, p.names, sp.names)
methods <- frug %>%
  group_by(Study_ID_Plus) %>%
  summarize(method = unique(Focus)) %>%
  mutate(method = ifelse(is.na(method), "Unknown", method))

for (ss in 1:nSP) {
  this_study  <- sp.names[ss]
  this_A      <- A_obs[, , ss]
  this_method <- methods[ss, "method"]
  these_v <- which(rowSums(this_A) > 0)
  these_p <- which(colSums(this_A) > 0)
  if (this_method == "Animal") {
    F_obs[these_v, , ss] <- 1
  } else if (this_method == "Plant") {
    F_obs[, these_p, ss] <- 1
  } else if (this_method == "BothLimited" | this_method == "Unknown") {
    F_obs[, , ss] <- this_A
  } else {
    F_obs[, , ss] <- 1
  }
}
```

## Sort species and load traits

```{r sort}
load(file.path(raw_path, "Cu_phylo.dat"))
load(file.path(raw_path, "Cv_phylo.dat"))
load(file.path(raw_path, "Obs_X.dat"))
load(file.path(raw_path, "Obs_W.dat"))
v.taxa <- read.csv(file.path(raw_path, "v_taxa_generic.csv"))
p.taxa <- read.csv(file.path(raw_path, "p_taxa_generic.csv"))

Cu <- Cu_phylo_generic
Cv <- Cv_phylo_generic
obs_W <- Obs_W_generic
obs_X <- Obs_X_generic

v.taxa.sorted <- v.taxa %>%
  arrange(Animal_Class, Animal_Family_Generic, Animal_Species_Generic)
v_species_order <- v.taxa.sorted$Animal_Species_Generic
Cu_sorted <- Cu[v_species_order, v_species_order]
obs_X <- obs_X[v_species_order, ]
O_V <- O_V[v_species_order, ]

p.taxa.sorted <- p.taxa %>%
  arrange(Plant_Family_Generic, Plant_Species_Generic)
p_species_order <- p.taxa.sorted$Plant_Species_Generic
Cv_sorted <- Cv[p_species_order, p_species_order]
obs_X <- obs_X[v_species_order, ]
O_P <- O_P[p_species_order, ]

obs_A <- A_obs[v_species_order, p_species_order, ]
obs_F <- F_obs[v_species_order, p_species_order, ]
```

## Save processed data

```{r save}
if (save_files) {
  save(obs_A,     file = file.path(processed_path, "A_obs.dat"))
  save(O_P,       file = file.path(processed_path, "OP_full.dat"))
  save(O_V,       file = file.path(processed_path, "OV_full.dat"))
  save(obs_F,     file = file.path(processed_path, "F_obs.dat"))
  save(Cu_sorted, file = file.path(processed_path, "Cu_phylo_sorted.dat"))
  save(Cv_sorted, file = file.path(processed_path, "Cv_phylo_sorted.dat"))
  save(obs_X,     file = file.path(processed_path, "Obs_X_sorted.dat"))
  save(obs_W,     file = file.path(processed_path, "Obs_W_sorted.dat"))
  write.csv(v.taxa.sorted, file.path(processed_path, "v_taxa_sorted.csv"), row.names = FALSE)
  write.csv(p.taxa.sorted, file.path(processed_path, "p_taxa_sorted.csv"), row.names = FALSE)
}
```
