---
title: "Fitting COIL+"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting COIL+}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

This vignette fits the COIL+ model to the Afrotropical frugivory data using
expert-defined occurrence probabilities and blocked sampling. Run
`vignette("01_data_prep")` first.

A small iteration count is used here for speed. Recommended settings for a
full analysis (Nsims = 500, burn = 10000, thin = 20) are noted in the code
comments.

```{r paths}
library(COILplus)
library(abind)
library(magrittr)

# Where processed data were saved by vignette("01_data_prep")
data_path <- file.path(system.file("extdata", package = "COILplus"), "processed")

date      <- "COILp_exp"
save_path <- file.path(tempdir(), "COILplus_results", date)
dir.create(save_path, recursive = TRUE, showWarnings = FALSE)
```

## Load processed data

```{r load}
load(file.path(data_path, "Cu_phylo_sorted.dat"))
load(file.path(data_path, "Cv_phylo_sorted.dat"))
load(file.path(data_path, "A_obs.dat"))
load(file.path(data_path, "F_obs.dat"))
load(file.path(data_path, "Obs_X_sorted.dat"))
load(file.path(data_path, "Obs_W_sorted.dat"))
load(file.path(data_path, "OP_full.dat"))
load(file.path(data_path, "OV_full.dat"))

Cu <- Cu_sorted
Cv <- Cv_sorted
```

## MCMC specifications

```{r specs}
# Reduced settings for illustration; increase for a real analysis
Nsims <- 10       # Recommended: 500
thin  <- 1        # Recommended: 20
burn  <- 5        # Recommended: 10000
use_H <- 10

sampling <- list(
  L = TRUE, lambda = TRUE, tau = TRUE, beta = TRUE,
  gamma = TRUE, sigmasq = TRUE, sigmasq_p = TRUE,
  delta = TRUE, zeta = TRUE, U = TRUE, V = TRUE, v = TRUE,
  z = TRUE, theta = TRUE, pis = TRUE, pjs = TRUE, rU = TRUE,
  rV = TRUE, miss_X = TRUE, miss_W = TRUE,
  O_V = TRUE, O_P = TRUE,   # sample occurrence indicators
  p_OV = TRUE, p_OP = TRUE  # sample occurrence probabilities
)
```

## Fit four chains

```{r fit}
n.chains <- 4
for (cc in 1:n.chains) {
  set.seed(cc)
  cat("\n Fitting chain", cc)

  mcmc <- MCMC(
    obs_A, focus = obs_F, p_occur_V = O_V, p_occur_P = O_P,
    obs_X, obs_W, Cu, Cv,
    Nsims, burn, thin, use_H = 10,
    use_shrinkage = TRUE, bias_cor = TRUE, theta_inf = 0.01,
    mh_n_pis = 100, mh_n_pjs = 100, mh_n_rho = 100,
    mh_pprior_sd = 0.1, mh_p_step = 0.1,
    stick_alpha = 5, prior_theta = c(1, 1), prior_tau = c(5, 5),
    prior_rho = c(5, 5), prior_mu0 = 0, prior_sigmasq0 = 10,
    prior_sigmasq = c(1, 1), start_values = NULL,
    sampling = sampling, cut_feed = FALSE, save_logL = TRUE
  )

  all_pred   <- abind::abind(pred_L = mcmc$Ls, probL = mcmc$mod_pL1s, along = 4)
  correlations <- cbind(U = mcmc$rU, V = mcmc$rV)
  p_detect   <- list(pis = mcmc$pi_mean, pjs = mcmc$pj_mean)
  factors    <- list(U = mcmc$U_mean, V = mcmc$V_mean)
  occ_plants <- list(OP_mean = mcmc$OP_mean, p_OPs = mcmc$p_OP_mean, p_accept = mcmc$p_OP_accepted)
  occ_verts  <- list(OV_mean = mcmc$OV_mean, p_VPs = mcmc$p_OV_mean, p_accept = mcmc$p_OV_accepted)

  res <- list(
    all_pred = all_pred, logL = mcmc$logL,
    correlations = correlations, p_detect = p_detect,
    factors = factors, occ_plants = occ_plants, occ_verts = occ_verts,
    Xs = mcmc$Xs, Ws = mcmc$Ws
  )
  save(res, file = file.path(save_path, paste0("res_", date, "_", cc, ".dat")))
  rm(res)
}
```
