---
title: "Cross-validation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cross-validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE)
```

This vignette performs k-fold cross-validation for the COIL+ model. Run
`vignette("01_data_prep")` first. Cross-validation for COIL without sampling occurrence probabilities will follow the same structure with the
`sampling` list adjusted accordingly (see `vignette("03_fit_coil")`).

A small iteration count is used here for speed. Recommended settings for a
full analysis (Nsims = 500, burn = 10000, thin = 20) are noted in the code
comments. Similarly, k = 10 is recommended, however we use a smaller number of folds for this vignette. 

```{r paths}
library(COILplus)
library(abind)

base_path      <- tools::R_user_dir("COILplus", which = "data")
raw_path       <- file.path(base_path, "raw")
processed_path <- file.path(base_path, "processed")
date      <- "COILp_exp_cv"
save_path <- file.path(base_path, "results", date)
dir.create(save_path, recursive = TRUE, showWarnings = FALSE)
```

```{r load}
load(file.path(processed_path, "Cu_phylo_sorted.dat"))
load(file.path(processed_path, "Cv_phylo_sorted.dat"))
load(file.path(processed_path, "A_obs.dat"))
load(file.path(processed_path, "F_obs.dat"))
load(file.path(processed_path, "Obs_X_sorted.dat"))
load(file.path(processed_path, "Obs_W_sorted.dat"))
load(file.path(processed_path, "OP_full.dat"))
load(file.path(processed_path, "OV_full.dat"))

Cu <- Cu_sorted
Cv <- Cv_sorted

nV <- nrow(obs_A)
nP <- ncol(obs_A)
nS <- dim(obs_A)[3]
```

```{r cv-setup}
Nsims <- 10      # Recommended: 500
thin  <- 1       # Recommended: 20
burn  <- 5       # Recommended: 10000
n.chains <- 4
n.folds  <- 2    # Recommended: k = 10

sampling <- list(
  L = TRUE, lambda = TRUE, tau = TRUE, beta = TRUE,
  gamma = TRUE, sigmasq = TRUE, sigmasq_p = TRUE,
  delta = TRUE, zeta = TRUE, U = TRUE, V = TRUE, v = TRUE,
  z = TRUE, theta = TRUE, pis = TRUE, pjs = TRUE, rU = TRUE,
  rV = TRUE, miss_X = TRUE, miss_W = TRUE,
  O_V = TRUE, O_P = TRUE, p_OV = TRUE, p_OP = TRUE
)

# Identify observed interactions and assign to folds
comb_A <- (apply(obs_A, c(1, 2), sum) > 0) * 1
obs_pairs <- which(comb_A == 1, arr.ind = TRUE)
set.seed(1)
fold_ids <- sample(rep(1:n.folds, length.out = nrow(obs_pairs)))
```

```{r cv-run}
for (kk in 1:n.folds) {
  cat("\n\n===== Fold", kk, "=====\n")

  # Mask held-out interactions
  held_out <- obs_pairs[fold_ids == kk, , drop = FALSE]
  use_A <- obs_A
  for (rr in seq_len(nrow(held_out))) {
    use_A[held_out[rr, 1], held_out[rr, 2], ] <- 0
  }

  for (cc in 1:n.chains) {
    set.seed(cc + kk * 100)

    mcmc <- MCMC(
      use_A, focus = obs_F, p_occur_V = O_V, p_occur_P = O_P,
      obs_X, obs_W, Cu, Cv,
      Nsims, burn, thin, use_H = 10,
      use_shrinkage = TRUE, bias_cor = TRUE, theta_inf = 0.01,
      mh_n_pis = 100, mh_n_pjs = 100, mh_n_rho = 100,
      mh_pprior_sd = 0.1, mh_p_step = 0.1,
      stick_alpha = 5, prior_theta = c(1, 1), prior_tau = c(5, 5),
      prior_rho = c(5, 5), prior_mu0 = 0, prior_sigmasq0 = 10,
      prior_sigmasq = c(1, 1), start_values = NULL,
      sampling = sampling, cut_feed = FALSE, save_logL = FALSE
    )

    all_pred <- abind::abind(pred_L = mcmc$Ls, probL = mcmc$mod_pL1s, along = 4)
    res <- list(all_pred = all_pred, held_out = held_out)
    save(res, file = file.path(save_path, paste0("cv_fold", kk, "_chain", cc, ".dat")))
    rm(res)
  }
}
```
