---
title: "Trait matching"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trait matching}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE)
```

This vignette estimates trait importance and direction (signed correlation)
for species interactions using permutation tests. Run
`vignette("02_fit_coilplus")` first.

A small permutation count is used here for speed. For a full analysis,
increase `B` to 1000 or more.

```{r packages}
library(COILplus)
library(tidyverse)
library(ggplot2)
library(grid)
library(gridExtra)
library(superheat)
library(abind)

base_path      <- tools::R_user_dir("COILplus", which = "data")
raw_path       <- file.path(base_path, "raw")
processed_path <- file.path(base_path, "processed")
date         <- "COILp_exp"
results_path <- file.path(base_path, "results", date)
save_path    <- results_path
```
## Load covariates and fitted interaction propensities. 
```{r load}
load(file.path(processed_path, "A_obs.dat"))
load(file.path(processed_path, "Obs_X_sorted.dat"))
load(file.path(processed_path, "Obs_W_sorted.dat"))

loadRData <- function(fileName) {
  load(fileName)
  get(ls()[ls() != "fileName"])
}

any_X_miss <- any(apply(obs_X, 2, function(x) sum(is.na(x))) > 0)
any_W_miss <- any(apply(obs_W, 2, function(x) sum(is.na(x))) > 0)

nchains  <- 4
all_pred <- vector("list", nchains)
all_X    <- vector("list", nchains)
all_W    <- vector("list", nchains)

for (ii in 1:nchains) {
  res <- loadRData(file.path(results_path, paste0("res_", date, "_", ii, ".dat")))
  all_pred[[ii]] <- res$all_pred
  if (any_X_miss) all_X[[ii]] <- res$Xs
  if (any_W_miss) all_W[[ii]] <- res$Ws
}
```

```{r combine}
library(abind)
mod_pL1s <- do.call(abind::abind, c(lapply(all_pred, function(x) x[, , , 2]), along = 1))

# Pool imputed covariates across chains
pool_imputed <- function(chain_list, ncov) {
  lapply(seq_len(ncov), function(mm) {
    do.call(rbind, lapply(chain_list, function(ch) ch[[mm]]))
  })
}

Nsims <- dim(mod_pL1s)[1]
pV <- if (any_X_miss) length(all_X[[1]]) else ncol(obs_X)
pP <- if (any_W_miss) length(all_W[[1]]) else ncol(obs_W)

Xs_pooled <- if (any_X_miss) pool_imputed(all_X, pV) else
  lapply(seq_len(pV), function(mm) matrix(NA, Nsims, 0))
Ws_pooled <- if (any_W_miss) pool_imputed(all_W, pP) else
  lapply(seq_len(pP), function(mm) matrix(NA, Nsims, 0))
```

## Run trait matching

Set `ncores > 1` to parallelise over posterior samples (recommended for large
datasets).

```{r trait-match}
# Sequential (ncores = 1) or parallel (ncores > 1)
ncores <- 1

trait_match <- TraitMatching(
  B          = 500,
  mod_pL1s   = mod_pL1s,
  Xs         = Xs_pooled,
  Ws         = Ws_pooled,
  obs_X      = obs_X,
  obs_W      = obs_W,
  obs_only   = FALSE,
  ncores     = ncores
)

save(trait_match, file = file.path(save_path, paste0("trait_match_", date, ".dat")))
```

```{r pvalues}
load(file.path(save_path, paste0("trait_match_", date, ".dat")))

# Proportion of permutation R-squared exceeding observed
pval_X <- sapply(seq_len(ncol(obs_X)), function(mm) {
  mean(trait_match$rsq_resampling_X[, mm] >= trait_match$rsq_obs_X[mm])
})
pval_W <- sapply(seq_len(ncol(obs_W)), function(mm) {
  mean(trait_match$rsq_resampling_W[, mm] >= trait_match$rsq_obs_W[mm])
})

results_df <- data.frame(
  covariate  = c(colnames(obs_X), colnames(obs_W)),
  species    = c(rep("Vertebrate", ncol(obs_X)), rep("Plant", ncol(obs_W))),
  rsq        = c(trait_match$rsq_obs_X, trait_match$rsq_obs_W),
  pvalue     = c(pval_X, pval_W)
)
print(results_df)
```

## Variable importance

Variable importance is measured as the observed R² for each trait minus the mean permuted R², expressed in standard deviations. Higher values indicate traits more predictive of interaction probability.

```{r plot-setup}
v.taxa <- read.csv(file.path(processed_path, "v_taxa_sorted.csv"))
p.taxa <- read.csv(file.path(processed_path, "p_taxa_sorted.csv"))
comb_A <- (apply(obs_A, c(1, 2), sum) > 0) * 1

good_namesX <- c("Generation Length", "Log Body Mass", "IUCN Status", "Habitat")
good_namesW <- c("Fruit Length", "Fruit Width", "Wood Density")

rsq_obs_X        <- trait_match$rsq_obs_X
rsq_obs_W        <- trait_match$rsq_obs_W
rsq_resampling_X <- trait_match$rsq_resampling_X
rsq_resampling_W <- trait_match$rsq_resampling_W
corr_obs_X       <- trait_match$corr_obs_X
corr_obs_W       <- trait_match$corr_obs_W

sd_awayX <- setNames(rep(NA, length(rsq_obs_X)), good_namesX)
for (cc in seq_along(rsq_obs_X))
  sd_awayX[cc] <- (rsq_obs_X[cc] - mean(rsq_resampling_X[, cc])) / sd(rsq_resampling_X[, cc])

sd_awayW <- setNames(rep(NA, length(rsq_obs_W)), good_namesW)
for (cc in seq_along(rsq_obs_W))
  sd_awayW[cc] <- (rsq_obs_W[cc] - mean(rsq_resampling_W[, cc])) / sd(rsq_resampling_W[, cc])
```
Here we compare the variable importance of each of the continuous frugivore covariates. 
```{r plot-importance-vert, fig.show='asis', fig.width=6, fig.height=4}
xx <- data.frame(value = sd_awayX, covariate = names(sd_awayX), y = 1)
xx <- xx[order(-xx$value), ]
xx$covariate <- factor(xx$covariate, levels = xx$covariate)
xx$value_rounded <- round(xx$value / 10) * 10

ggplot() + geom_tile(aes(x = covariate, y = y, fill = value_rounded), color = 'white', data = xx) +
  theme_bw() + theme(panel.border = element_blank(), panel.grid = element_blank()) +
  theme(axis.text.y = element_blank(), axis.ticks = element_blank()) +
  theme(legend.position = 'bottom', axis.title = element_blank()) +
  guides(fill = guide_legend("Frugivore Variable Importance")) +
  scale_fill_gradient(low = '#BFF0B6', high = '#3B6E32', breaks = xx$value_rounded) +
  theme(axis.text = element_text(angle = 0, hjust = 0.5, vjust = 0, size = 20, family = "serif", color = "black"),
        legend.text = element_text(angle = 0, hjust = 0.5, vjust = 0, size = 15, family = "serif", color = "black"),
        legend.title = element_text(size = 20, family = "serif"))
```

And here we compare the variable importance of each of the plant covariates. 
```{r plot-importance-plant, fig.show='asis', fig.width=6, fig.height=4}
ww <- data.frame(value = sd_awayW, covariate = names(sd_awayW), y = 1)
ww <- ww[order(-ww$value), ]
ww$covariate <- factor(ww$covariate, levels = ww$covariate)
ww$value_rounded <- round(ww$value / 10) * 10

ggplot() + geom_tile(aes(x = covariate, y = y, fill = value_rounded), color = 'white', data = ww) +
  theme_bw() + theme(panel.border = element_blank(), panel.grid = element_blank()) +
  theme(axis.text.y = element_blank(), axis.ticks = element_blank()) +
  theme(legend.position = 'bottom', axis.title = element_blank()) +
  scale_fill_gradientn(colors = c('#D4DEF7', '#425075'), breaks = ww$value_rounded) +
  guides(fill = guide_legend("Plant Variable Importance")) +
  theme(axis.text = element_text(angle = 0, hjust = 0.5, vjust = 0, size = 20, family = "serif", color = "black"),
        legend.text = element_text(angle = 0, hjust = 0.5, vjust = 0, size = 15, family = "serif", color = "black"),
        legend.title = element_text(size = 20, family = "serif"))
```

## Posterior interactions by trait rank

Species are ordered by their value on the top-ranked trait. Observed interactions (black) are masked to highlight predicted novel interactions.

```{r plot-trait-raster, fig.show='asis', fig.width=8, fig.height=6}
use_Ls      <- do.call(abind::abind, c(lapply(all_pred, function(x) x[, , , 1]), along = 1))
use_mean_Ls <- apply(use_Ls, c(2, 3), mean)
use_mean_Ls[comb_A == 1] <- NA

wh_X <- 1
wh_W <- 1
keep_birds  <- which(!is.na(obs_X[, wh_X]))
keep_plants <- which(!is.na(obs_W[, wh_W]))
use_out <- use_mean_Ls[keep_birds, keep_plants]

bird_cov  <- obs_X[keep_birds, wh_X]
bird_cov  <- bird_cov + rnorm(length(keep_birds), sd = sd(bird_cov) * 0.0001)
plant_cov <- obs_W[keep_plants, wh_W]
plant_cov <- plant_cov + rnorm(length(plant_cov), sd = sd(plant_cov) * 0.0001)

plot_dta <- data.frame(cov_bird    = rep(bird_cov, length(keep_plants)),
                       cov_plant   = rep(plant_cov, each = length(keep_birds)),
                       probability = as.numeric(use_out))
plot_dta$use_cov_bird  <- factor(as.numeric(factor(plot_dta$cov_bird)))
plot_dta$use_cov_plant <- factor(as.numeric(factor(plot_dta$cov_plant)))

g <- ggplot() +
  geom_raster(aes(x = use_cov_plant, y = use_cov_bird, fill = probability), data = plot_dta) +
  scale_fill_gradient(low = "#FDE725FF", high = "#02A65F", na.value = '#016B3B',
                      name = 'Posterior\ninteraction\nprobability\n', limits = c(0, 1)) +
  theme(axis.ticks = element_blank(), axis.text = element_blank(),
        axis.title = element_text(size = 30),
        axis.title.y = element_text(vjust = -1, family = 'serif'),
        text = element_text(family = 'serif')) +
  ylab(expression(symbol('\256'))) + xlab(expression(symbol('\256')))

gridExtra::grid.arrange(g,
  left   = textGrob("Frugivores: increasing log body mass", rot = 90,
                    x = 1.3, y = 0.57, gp = gpar(fontsize = 12, fontfamily = 'serif')),
  bottom = textGrob("Plants: increasing wood density",
                    x = 0.435, y = 1.3, gp = gpar(fontsize = 12, fontfamily = 'serif')),
  vp = viewport(width = 0.5, height = 0.6))
```

## Signed correlations by taxonomic group
Now we explore heterogeneity in trait-interaction relationships by look at the signed correlation across families. Beginning with plant traits, we look at the signed correlation between a trait and interaction probability for a given frugivore family. Families with fewer than the minimum size threshold are excluded.

```{r plot-signed-corr-plant-traits, fig.show='asis', fig.width=8, fig.height=8}
v_group <- v.taxa$Animal_Family_Generic
p_group <- p.taxa$Plant_Family_Generic

v_size_cluster <- sapply(unique(v_group), function(x) sum(v_group == x))
p_size_cluster <- sapply(unique(p_group), function(x) sum(p_group == x))

keep_v_groups <- sort(names(which(v_size_cluster >= 12)))
keep_p_groups <- names(which(p_size_cluster >= 30))
keep_v_index  <- which(v_group %in% keep_v_groups)
keep_p_index  <- which(p_group %in% keep_p_groups)

corr_obs_W_plot <- corr_obs_W
colnames(corr_obs_W_plot) <- c("Fruit length", "Fruit width", "Wood density")

superheat(X = corr_obs_W_plot[keep_v_index, ],
          pretty.order.cols = FALSE, pretty.order.rows = FALSE,
          membership.rows = v_group[keep_v_index],
          grid.hline.col = "black", grid.vline.col = 'black',
          grid.hline.size = 0.5, grid.vline.size = 0.5,
          left.label = "none", legend = TRUE,
          bottom.label.text.size = 10, bottom.label.size = 0.075)
```
Now we do the same for frugivore traits, evaluating how their association with interaciton propensity varies across plant families. 
```{r plot-signed-corr-vert-traits, fig.show='asis', fig.width=8, fig.height=8}
plot_pred_X <- corr_obs_X[, -c(3, 4)]
colnames(plot_pred_X) <- c("Generation length", "Log body mass")

superheat(X = plot_pred_X[keep_p_index, ],
          pretty.order.cols = FALSE, pretty.order.rows = FALSE,
          membership.rows = p_group[keep_p_index],
          left.label = "none",
          grid.hline.col = "black", grid.vline.col = 'black',
          legend = TRUE,
          bottom.label.text.size = 10, bottom.label.size = 0.075)
```
