---
title: "Trait matching"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trait matching}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

This vignette estimates trait importance and direction (signed correlation)
for species interactions using permutation tests. Run
`vignette("02_fit_coilplus")` first.

A small permutation count is used here for speed. For a full analysis,
increase `B` to 1000 or more.

```{r packages}
library(COILplus)
library(tidyverse)
library(grid)
library(gridExtra)

data_path    <- file.path(system.file("extdata", package = "COILplus"), "processed")
date         <- "COILp_exp"
results_path <- file.path(tempdir(), "COILplus_results", date)
save_path    <- results_path
```

```{r load}
load(file.path(data_path, "A_obs.dat"))
load(file.path(data_path, "Obs_X_sorted.dat"))
load(file.path(data_path, "Obs_W_sorted.dat"))

loadRData <- function(fileName) {
  load(fileName)
  get(ls()[ls() != "fileName"])
}

any_X_miss <- any(apply(obs_X, 2, function(x) sum(is.na(x))) > 0)
any_W_miss <- any(apply(obs_W, 2, function(x) sum(is.na(x))) > 0)

nchains  <- 4
all_pred <- vector("list", nchains)
all_X    <- vector("list", nchains)
all_W    <- vector("list", nchains)

for (ii in 1:nchains) {
  res <- loadRData(file.path(results_path, paste0("res_", date, "_", ii, ".dat")))
  all_pred[[ii]] <- res$all_pred
  if (any_X_miss) all_X[[ii]] <- res$Xs
  if (any_W_miss) all_W[[ii]] <- res$Ws
}
```

```{r combine}
library(abind)
mod_pL1s <- do.call(abind::abind, c(lapply(all_pred, function(x) x[, , , 2]), along = 1))

# Pool imputed covariates across chains
pool_imputed <- function(chain_list, ncov) {
  lapply(seq_len(ncov), function(mm) {
    do.call(rbind, lapply(chain_list, function(ch) ch[[mm]]))
  })
}

Nsims <- dim(mod_pL1s)[1]
pV <- if (any_X_miss) length(all_X[[1]]) else ncol(obs_X)
pP <- if (any_W_miss) length(all_W[[1]]) else ncol(obs_W)

Xs_pooled <- if (any_X_miss) pool_imputed(all_X, pV) else
  lapply(seq_len(pV), function(mm) matrix(NA, Nsims, 0))
Ws_pooled <- if (any_W_miss) pool_imputed(all_W, pP) else
  lapply(seq_len(pP), function(mm) matrix(NA, Nsims, 0))
```

## Run trait matching

Set `ncores > 1` to parallelise over posterior samples (recommended for large
datasets).

```{r trait-match}
# Sequential (ncores = 1) or parallel (ncores > 1)
ncores <- 1

trait_match <- TraitMatching(
  B          = 500,
  mod_pL1s   = mod_pL1s,
  Xs         = Xs_pooled,
  Ws         = Ws_pooled,
  obs_X      = obs_X,
  obs_W      = obs_W,
  obs_only   = FALSE,
  ncores     = ncores
)

save(trait_match, file = file.path(save_path, paste0("trait_match_", date, ".dat")))
```

## Compute p-values

```{r pvalues}
load(file.path(save_path, paste0("trait_match_", date, ".dat")))

# Proportion of permutation R-squared exceeding observed
pval_X <- sapply(seq_len(ncol(obs_X)), function(mm) {
  mean(trait_match$rsq_resampling_X[, mm] >= trait_match$rsq_obs_X[mm])
})
pval_W <- sapply(seq_len(ncol(obs_W)), function(mm) {
  mean(trait_match$rsq_resampling_W[, mm] >= trait_match$rsq_obs_W[mm])
})

results_df <- data.frame(
  covariate  = c(colnames(obs_X), colnames(obs_W)),
  species    = c(rep("Vertebrate", ncol(obs_X)), rep("Plant", ncol(obs_W))),
  rsq        = c(trait_match$rsq_obs_X, trait_match$rsq_obs_W),
  pvalue     = c(pval_X, pval_W)
)
print(results_df)
```

## Plot signed correlations

```{r plot}
# See Analysis_4b_trait_matching_plot.R for full plotting code
# Signed direction: trait_match$corr_obs_X (vertebrate covariates x plant species)
#                   trait_match$corr_obs_W (plant covariates x vertebrate species)
```
